---
title: "Sequential testing of hypotheses about population density with the `sequential.pops` package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Seq-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `sequential.pops` package is designed to simplify the development, evaluation 
and implementation of sequential designs for testing hypotheses about population
density. Sequential analyses are particularly useful for decision-making in situations 
where estimation is not the goal, but when critical actions should be triggered if
population sizes reach or are above or below predefined levels. This situations include
deciding whether a pest control action should be applied, whether an area should be
declared free from an invasive species, or whether early capture numbers of a pest
are consistent with an outbreak. Sequential analyses can also be useful in fields 
outside ecology and pest management, such as quality control, and adaptive clinical 
trials, but those applications are out of the scope of this package.


```{r setup}
library(sequential.pops)
```
