---
title: "Sequential testing of hypotheses about population density with the `sequential.pops` package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Seq-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `sequential.pops` package is designed to simplify the development, evaluation 
and analysis of sequential designs for testing hypotheses about population
density. Sequential analyses are particularly useful for decision-making in situations 
where estimation is not the goal, but where critical actions should be triggered if
population sizes reach or are above or below predefined levels. This situations include
deciding whether a pest control action should be applied, whether an area should be
declared free from an invasive species, or whether early capture numbers of a pest
are consistent with an outbreak. Sequential analyses can also be useful in fields 
outside ecology and pest management, such as quality control, and adaptive clinical 
trials, but those applications are out of the scope of this package.

Sequential data come in sampling bouts over time that should be processed as they come 
to evaluate one or several hypotheses and stop collecting data (sampling) when there
is enough evidence to make a decision. Sequential designs are typically more cost-efficient
than conventional fixed-sample-size approaches, and can be purely sequential (one-at-a-time)
or group sequential (`n > 1` per sampling bout). The analysis of biological population
densities is often carried out by collecting count data from traps or field observations
or binomial records from structured clusters of sampling units. The `sequential.pops`
package focuses on count and binomial data with or without overdispersion to test hypotheses
about non-negative population sizes.

There are other R packages that deal with sequential designs. Most existing
packages, such as `GroupSeq`, `Grpseq`, `gsbDesign` and `seqmon`, focus on adaptive clinical 
trials for normally distributed variables. Others, such as `gscounts`, `gsDesign`, `ASSISTant`,
`Sequential` and `BayesOrdDesign` can take non-normal variables but are also focused 
on clinical trials and can hardly be adapted to process count or binomial data from 
ecological or agricultural settings and evaluate hypotheses about population densities. 
Other packages with a more general scope, such as `MSPRT` and `sprtt`, include a limited 
selection of probability distributions and lack functions to evaluate tests' operating
characteristics.

The `sequential.pops` is the first R package devoted to test hypotheses on biological 
population densities. Two approaches are included: the Sequential Test of Bayesian
Posterior Probabilities (STBP) (Rincon et al. 2025), and the Sequential Probability
Ratio Test (SPRT) (Wald 1945). Both the STBP and the SPRT have their own pros and cons 
but are the most popular and state-of-the-art approaches for sequential analyses in 
ecology and pest management. This tutorial walks you through the development, evaluation 
and analysis of sequential designs with STBP and SPRT using the `sequential.pops`
package.

## 1. The Sequential Test of Bayesian Posterior Probabilities

This test is specifically designed to test hypotheses about population densities,
although it may have applications in other fields. It works by sequentially updating 
the conditional probability of the tested hypothesis as new data is processed. The
two main benefits of using STBP is that is distribution-agnostic (works with virtually
any probability distribution), and that is one of the few sequential analyses that
is "likelihood ratio-free" meaning that it can test single hypotheses without having 
to specify a non-complementary alternative. As we'll see later, most sequential tests, 
such as the SPRT (section 2), require two hypotheses, rather than one, because they are based on 
likelihood ratios. Another advantage of the STBP is that it can process purely 
sequential (one-at-a-time) or group sequential (in batches of samples), balanced or 
unbalanced, and static (single-value) or dynamic (population models) hypotheses without 
major modifications. 

The main disadvantage of the STBP is that it is so new that has not been as tested as its
counterparts in the rough streets of real datasets. The STBP is also pretty sensitive to
sample sizes within sampling bouts. In theory, it outperforms other sequential analysis
as long as several samples are collected within each sampling bout, but type I error rate
(accepting H when is false) gets ugly in purely sequential designs.

### Example 1: Testing species absence

Let's start with a simple case in which one wants to test the hypothesis that certain
species is absent in sampled area. As Carl Sagan said "the absence of evidence is not 
evidence of absence". In other words, we can never be sure about the absence of an species
in an area, but we can calculate the *probability* of the species being absent *given* 
the collected data.


```{r setup}
library(sequential.pops)
```
