---
title: "Sequential testing of hypotheses about population density with the `sequential.pops` package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Seq-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `sequential.pops` package is designed to simplify the development, evaluation 
and implementation of sequential designs for testing hypotheses about population
density. Sequential analyses are particularly useful for decision-making in situations 
where estimation is not the goal, but where critical actions should be triggered if
population sizes reach or are above or below predefined levels. This situations include
deciding whether a pest control action should be applied, whether an area should be
declared free from an invasive species, or whether early capture numbers of a pest
are consistent with an outbreak. Sequential analyses can also be useful in fields 
outside ecology and pest management, such as quality control, and adaptive clinical 
trials, but those applications are out of the scope of this package.

Sequential data come in sampling bouts over time that should be processed as they come 
to evaluate one or several hypotheses and stop collecting data (sampling) when there
is enough evidence to make a decision. Sequential designs are typically more cost-efficient
than conventional fixed-sample-size approaches, and can be purely sequential (one-at-a-time)
or group sequential (`n > 1` per sampling bout). The analysis of biological population
densities is often carried out by collecting count data from traps or field observations
or binomial records from structured clusters of sampling units. The `sequential.pops`
package focuses on count and binomial data with or without overdispersion to test hypotheses
about non-negative population sizes.

There are other R packages that deal with sequential designs. Most existing
packages, such as `GroupSeq`, `Grpseq`, `gsbDesign` and `seqmon`, focus on adaptive clinical 
trials for normally distributed variables. Others, such as `gscounts`, `gsDesign`, `ASSISTant`,
`Sequential` and `BayesOrdDesign` can take non-normal variables but are also focused 
on clinical trials and can hardly be adapted to process count or binomial data from 
ecological or agricultural settings and evaluate hypotheses about population densities. 
Other packages with a more general scope, such as `MSPRT` and `sprtt`, include a limited 
selection of probability distributions and lack functions to evaluate test from operating
characteristics.

The `sequential.pops` is the first R package devoted to test hypotheses on biological 
population densities. Two approaches are included: the Sequential Test of Bayesian
Posterior Probabilities (STBP) (Rincon et al. 2025), and the Sequential Probability
Ratio Test (SPRT) (Wald 1945). Both the STBP and the SPRT have their own pros and cons 
but are the most popular and state-of-the-art approaches for sequential analyses in 
ecology and pest management. This tutorial walks you through the development, evaluation
and implementation of sequential designs with STBP and SPRT using the `sequential.pops`
package.

## The Sequential Test of Bayesian Posterior Probabilities



```{r setup}
library(sequential.pops)
```
